Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument

### 2014-10-17 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
update of early draft  paper

%----------AUTHOR 1
\author[Benzm\"uller]{Christoph Benzm\"uller}

@@

\thanks{This work was completed with the support of ToDo}

%----------Author 2
\author[Weber]{\\ Leon Weber}
\address{ 
ToDo \\
ToDo \\
ToDo Berlin \\
Germany
}
\email{leon.weber@fu-berlin.de}

%----------Author 3
\author[Woltzenlogel-Paleo]{\\ Bruno Woltzenlogel Paleo}
\address{ 
Favoritenstra{\ss}e 9 \\
Room HA0402 \\
1040 Wien \\
Austria
}
\email{bruno.wp@gmail.com}

@@

\section{The Modal Collapse}
The axioms in G\"odel's ontological argument \cite{GoedelNotes,ScottNotes} imply what is
called the modal collapse \cite{Sobel1987,SobelBook} (formula $\varphi \rightarrow \Box \varphi$
holds for any formula $\varphi$ and not just for $\exists x. God(x)$
as intended). This fact, which has recently been confirmed with
higher-order automated theorem provers \cite{}, has led to strong
criticism in the research community, and it has stimulated further
adaptations of G\''odel's argument with the aim to remedy the situation
\cite{}.

In recent work we have formalized and automated the proposed adaptions
of Anderson \cite{}, Bjordahl \cite{} and Fuhrmann \cite{}.

\pagebreak

 crucial step of the ontological argument is the claim that if God's existence is possible, then it is necessary. This is Lemma L2 in G\"odel's proof. In the modal square of opposition (Fig. \ref{ToDo}), this is an unusual situation in which the \textbf{I} corner implies and entails the \textbf{A} corner, for the particular case when $\phi$ is $\exq x G(x)$. G\"odel's proof shows that his axioms are strong enough to invert the direction of entailment for the particular $\phi$ at issue. The question is whether the axioms are then not too strong, also allowing the inverted entailment for arbitrary $\phi$. That is essentially the question asked by Sobel \cite{Sobel1987}; and his proof of the modal collapse (MC) answers it affirmatively. The modal collapse is a collapse of the modal square: it shows that, assuming G\"odel's axioms, the subcontraries entail the contraries.

ToDo: include figure of Modal Square of Opposition.

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[MC] Everything that is the case is so necessarily:
  $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
\item[MC'] Everything that is possible is necessary:
  $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
\item[MC''] Modalities collapse:
  $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
\end{figure}

\clearpage

\section{Anderson's Emendation}

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A:A1] If a property is positive, its negation is not positive:
  $$\hol{\allq \varphi [P(\varphi) \imp \neg P(\neg \varphi)]}$$ 
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \varphi \allq \psi [(P(\varphi) \wedge \nec \allq x [\varphi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified: 
  $$\hol{\allq \varphi [P(\varphi) \imp \pos \exq x \varphi(x)]}$$ 
\item[A:D1] A \emph{God-like} being necessarily possesses those and only those properties that are positive: 
  $$\hol{G_A(x) \equiv \forall \varphi [P(\varphi) \biimp \nec \varphi(x)]}$$ 
\item[A3']  The property of being God-like is positive: 
  $$\hol{P(G_A)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive: 
  $$\hol{\allq \varphi [P(\varphi) \imp \Box \; P(\varphi)]}$$ 
\item[A:D2] An \emph{essence} of an individual is a property that necessarily implies those and only those properties that the individual has necessarily: $$\hol{\essA{\varphi}{x} \equiv \allq
  \psi [\nec \psi(x) \biimp \nec \allq y (\varphi(y) \imp \psi(y))]}$$ 
\item[T2']  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G_A(x) \imp \essA{G_A}{x}]}$$
\item[D3'] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE_A(x) \equiv \allq \varphi [\essA{\varphi}{x} \imp \nec
  \exq y \varphi(y)]}$$
\item[A5'] Necessary existence is a positive property: $$\hol{P(\NE_A)}$$ 
\item[L1'] If a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\exq x G_A(x) \imp \nec \exq y G_A(y)}$$
\item[L2'] If possibly a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\pos \exq x G_A(x) \imp \nec \exq y G_A(y)} $$
%
\item[T3'] Necessarily, a God-like being exists: $$\hol{\nec \exq x G_A(x)}$$ 
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Anderson's Emendation \cite{ToDo:Anderson}.\label{fig:anderson}} 
\end{figure}

I have meanwhile added another file ``Anderson\_var\_partial.thy''. In this version the actualist e-Quantifier is used exclusively in the definition of NE and everywhere else ithe possibilist Quantifiers are used. It is still unclear to me, what is actually meant. Reading Anderson 1990, last sentence on p. 302, it could mean that ``Anderson\_var\_partial.thy'' is the correct version; but when reading AndersonGettings, then one could think that ``Anderson\_var.thy'' is meant. But ok we have now both versions. Interestingly in ``Anderson\_var\_partial.thy'' A5 is not inferable anymore, and this also holds for ``anderson\_implies\_fuhrmann". Nitpick finds countermodels.

\clearpage

\section{Frode's Alternative}

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A1] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \phi [P(\neg \phi) \biimp \neg P(\phi)]}$$ 
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \phi \allq \psi [(P(\phi) \wedge \nec \allq x [\phi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified: 
  $$\hol{\allq \phi [P(\phi) \imp \pos \exq x \phi(x)]}$$ 
\item[D1] A \emph{God-like} being possesses all positive properties: 
  $$\hol{G(x) \biimp \forall \phi [P(\phi) \imp \phi(x)]}$$ 
\item[A3]  The property of being God-like is positive: 
  $$\hol{P(G)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive: 
  $$\hol{\allq \phi [P(\phi) \imp \Box \; P(\phi)]}$$ 
\item[D2] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\phi}{x} \biimp \phi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\phi(y) \imp \psi(y)))}$$ 
\item[T2]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[D3] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE(x) \biimp \allq \phi [\ess{\phi}{x} \imp \nec
  \exq y \phi(y)]}$$
\item[A5] Necessary existence is a positive property: $$\hol{P(\NE)}$$ 
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Frode's Alternative \cite{ToDo:Frode}.\label{fig:frode}} 
\end{figure}

\clearpage

\section{Other Solutions}

ToDo: Fitting

\section{Conclusions}

Various slightly different versions of
axioms and definitions have been considered by G\"{o}del and by
several philosophers who commented on his proof
(cf.~\cite{sobel2004logic,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).

In theoretical philosophy, formal logical confrontations with such
ontological arguments had been so far (mainly) limited to paper
and pen.  Up to now, the use of computers was prevented, because the
logics of the available theorem proving systems were not expressive
enough to formalize the abstract concepts adequately. G{\"o}del's proof
uses, for example, a complex higher-order modal logic (\HOML) to handle
concepts such as \emph{possibility} and \emph{necessity} and to support
quantification over individuals and properties.

controversies, care with parameters

ToDo: Leibniz calculemus, Rushby, Zalta \cite{oppenheimera11,rushby13}

@@

>>>>

%----------AUTHOR 1
\author[Benzm\"uller]{C. Benzm\"uller}

@@

\thanks{This work was supported by German National Research Foundation (DFG) under
 grants BE 2501/9-1 and BE 2501/11-1 and by ????}

%----------Author 2
\author[Weber]{L. Weber}
\address{ 
ToDo \\
ToDo \\
ToDo Berlin \\
Germany
}
\email{leon.weber@fu-berlin.de}

%----------Author 3
\author[Woltzenlogel-Paleo]{B. Woltzenlogel Paleo}
\address{ 
Favoritenstra{\ss}e 9 \\
Room HA0402 \\
1040 Wien \\
Austria
}
\email{bruno.wp@gmail.com}

@@

%\vspace*{-5em}

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument
\cite{GoedelNotes,ScottNotes} imply what is called the modal collapse
\cite{Sobel1987,SobelBook} (formula $\varphi \rightarrow \Box \varphi$
holds for any formula $\varphi$ and not just for $\exists x. God(x)$
as intended). This fact, which has recently been confirmed with
higher-order automated theorem provers \cite{J30,ECAI-2014}, has led
to strong criticism of the argument, and it has therefore stimulated
attempts to remedy the problem. Some authors, including Hajek
\cite{Hajek_der_Mathematiker,Hajek_Magari_and_others}, suggest the use
of cautious instead of full comprehension principles in order to
escape the modal collapse. Others, including Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordahl \cite{}, have proposed  to adapt G\"odel's axioms (in
particular, controversial axiom A1) and definitions.

We have studied --- with computer-support --- the emendations proposed
by Anderson and Bjordahl.  More precisely, we have formalized their
emendations within the proof assistant Isabelle/HOL \cite{Isabelle},
and we have then automated (different variations of) their
emendations with the higher-order reasoners Leo-II \cite{}, Satallax
\cite{} and Nitpick \cite{} available in Isabelle/HOL.\footnote{The
  formalizations are available in the subdirectories \url{Anderson}
  and \url{Bjordahl} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
Our formalizations employ the embedding of higher-order modal logic
(HOML) in classical higher-order logic (HOL) as introduced and
employed in our previous work \cite{J30,ECAI-2014}. In this approach
full comprehension is naturally ``built-in'' since the underlying HOL
supports $\lambda$-abstraction. We summarize the main results of our
experiments.

Anderson's emendation (cf. Appendix \ref{apx:Anderson}) has been
verified for different domain conditions. These variations were
motivated by different authors who commented on Anderson's work. Anderson himself \ldots
\begin{itemize}
\item Constant domain semantics \footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_constant_domain.thy}}, that is, the domains of
  objects for all types (individuals, predicates, etc.) stays the same
  in all worlds. We consider constant domain semantics as the standard
  setting.
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source file \url{Anderson/Anderson_varying_domain.thy}} for the type of indivduals individuals
  (and constant domain for all other types). Different authors have pointed out that 
  Ande
\item The foll
and constant 
has been verified for constant domain\footnote{Cf. Isabelle/HOL source file \url{Anderson/Anderson_const.thy}} and varying
  domain semantics
\item 
\end{itemize}

 that is, the following can be proved by the HOL-ATPs:

G\"odel's and Scott's versions of the proof are not explicit about the
particular comprehension principles they assume.  Some authors,
including Hajek \cite{Hajek_der_Mathematiker,Hajek_Magari_and_others},
have suggested to consider cautious conprehension principles in order
to escape the modal collapse.

% \section{Other Solutions}

% ToDo: Fitting

% \section{Conclusions}

% Various slightly different versions of
% axioms and definitions have been considered by G\"{o}del and by
% several philosophers who commented on his proof
% (cf.~\cite{sobel2004logic,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).

% In theoretical philosophy, formal logical confrontations with such
% ontological arguments had been so far (mainly) limited to paper
% and pen.  Up to now, the use of computers was prevented, because the
% logics of the available theorem proving systems were not expressive
% enough to formalize the abstract concepts adequately. G{\"o}del's proof
% uses, for example, a complex higher-order modal logic (\HOML) to handle
% concepts such as \emph{possibility} and \emph{necessity} and to support
% quantification over individuals and properties.

% controversies, care with parameters

% ToDo: Leibniz calculemus, Rushby, Zalta \cite{oppenheimera11,rushby13}

@@

\begin{appendix}

% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
% \begin{itemize}
% \item[MC] Everything that is the case is so necessarily:
%   $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
% \item[MC'] Everything that is possible is necessary:
%   $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
% \item[MC''] Modalities collapse:
%   $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
% \end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
% \end{figure}

%\clearpage

\section{Anderson's Emendation} \label{apx:Anderson}

% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A:A1] If a property is positive, its negation is not positive:
  $$\hol{\allq \varphi [P(\varphi) \imp \neg P(\neg \varphi)]}$$ 
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \varphi \allq \psi [(P(\varphi) \wedge \nec \allq x [\varphi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified: 
  $$\hol{\allq \varphi [P(\varphi) \imp \pos \exq x \varphi(x)]}$$ 
\item[A:D1] A \emph{God-like} being necessarily possesses those and only those properties that are positive: 
  $$\hol{G_A(x) \equiv \forall \varphi [P(\varphi) \biimp \nec \varphi(x)]}$$ 
\item[A3']  The property of being God-like is positive: 
  $$\hol{P(G_A)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive: 
  $$\hol{\allq \varphi [P(\varphi) \imp \Box \; P(\varphi)]}$$ 
\item[A:D2] An \emph{essence} of an individual is a property that necessarily implies those and only those properties that the individual has necessarily: $$\hol{\essA{\varphi}{x} \equiv \allq
  \psi [\nec \psi(x) \biimp \nec \allq y (\varphi(y) \imp \psi(y))]}$$ 
\item[T2']  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G_A(x) \imp \essA{G_A}{x}]}$$
\item[D3'] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE_A(x) \equiv \allq \varphi [\essA{\varphi}{x} \imp \nec
  \exq y \varphi(y)]}$$
\item[A5'] Necessary existence is a positive property: $$\hol{P(\NE_A)}$$ 
\item[L1'] If a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\exq x G_A(x) \imp \nec \exq y G_A(y)}$$
\item[L2'] If possibly a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\pos \exq x G_A(x) \imp \nec \exq y G_A(y)} $$
%
\item[T3'] Necessarily, a God-like being exists: $$\hol{\nec \exq x G_A(x)}$$ 
\end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Anderson's Emendation \cite{ToDo:Anderson}.\label{fig:anderson}} 
% \end{figure}

% I have meanwhile added another file ``Anderson\_var\_partial.thy''. In this version the actualist e-Quantifier is used exclusively in the definition of NE and everywhere else ithe possibilist Quantifiers are used. It is still unclear to me, what is actually meant. Reading Anderson 1990, last sentence on p. 302, it could mean that ``Anderson\_var\_partial.thy'' is the correct version; but when reading AndersonGettings, then one could think that ``Anderson\_var.thy'' is meant. But ok we have now both versions. Interestingly in ``Anderson\_var\_partial.thy'' A5 is not inferable anymore, and this also holds for ``anderson\_implies\_fuhrmann". Nitpick finds countermodels.

%\clearpage

\section{Bjordahl's Alternative} \label{apx:Bjordahl}

% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A1] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \phi [P(\neg \phi) \biimp \neg P(\phi)]}$$ 
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \phi \allq \psi [(P(\phi) \wedge \nec \allq x [\phi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified: 
  $$\hol{\allq \phi [P(\phi) \imp \pos \exq x \phi(x)]}$$ 
\item[D1] A \emph{God-like} being possesses all positive properties: 
  $$\hol{G(x) \biimp \forall \phi [P(\phi) \imp \phi(x)]}$$ 
\item[A3]  The property of being God-like is positive: 
  $$\hol{P(G)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive: 
  $$\hol{\allq \phi [P(\phi) \imp \Box \; P(\phi)]}$$ 
\item[D2] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\phi}{x} \biimp \phi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\phi(y) \imp \psi(y)))}$$ 
\item[T2]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[D3] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE(x) \biimp \allq \phi [\ess{\phi}{x} \imp \nec
  \exq y \phi(y)]}$$
\item[A5] Necessary existence is a positive property: $$\hol{P(\NE)}$$ 
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Frode's Alternative \cite{ToDo:Frode}.\label{fig:frode}} 
% \end{figure}

%\clearpage
\end{appendix}
